# Defining your own functions

In this section we are going to learn some advanced concepts that are going to make you into a
full-fledged R programmer. Before this chapter you only used whatever R came with, as well as the
R packages. This already allowed you to do a lot of things and solve a variety of problems!
The next step is now to learn how to actually build your own functions.

## Control flow

Knowing about control flow is essential to build your own functions. Without control flow statements,
such as if-else statements or loops (or, in the case of pure functional programming languages, recursion),
programming languages would be very limited.

### If-else

Imagine you want a variable to be equal to a certain value if a condition is met. This is a typical
problem that requires the `if ... else ...` construct. For instance:

```{r, cache=TRUE}
a <- 4
b <- 5
```

Suppose that if `a > b` then `f` should be equal to 20, else `f` should be equal to 10. Using ```if ... else
...``` you can achieve this like so:

```{r, cache=TRUE}
if (a > b) {
  f <- 20
    } else {
  f <- 10
}
```

Obviously, here `f = 10`. Another way to achieve this is by using the `ifelse()` function:

```{r, cache=TRUE}
f <- ifelse(a > b, 20, 10)
```

`if...else...` and `ifelse()` might seem interchangeable, but they're not. `ifelse()` is vectorized, while
`if...else..` is not. Let's try the following:

```{r}
ifelse(c(1,2,4) > c(3, 1, 0), "yes", "no")
```

The result is a vector. Now, let's see what happens if we use `if...else...` instead of `ifelse()`:

```{r}
if (c(1, 2, 4) > c(3, 1, 0)) print("yes") else print("no")
```

Only the first element of my atomic vector is used for the comparison. This is very important to keep in mind.
Suppose that you want an expression to be evaluated, only if every element is `TRUE`. In this case, you should
use the `all()` function, as seen previously in Chapter 2:

```{r}
ifelse(all(c(1,2,4) > c(3, 1, 0)), "all elements are greater", "not all elements are greater")
```

You also remember the `any()` function:

```{r}
ifelse(any(c(1,2,4) > c(3, 1, 0)), "at least one element is greater", "no element greater")
```

These are the basics. But sometimes, you might need to test for more complex conditions, which can lead to using
nested `if...else...` constructs. These, however, can get messy:

```{r, cache=TRUE}
if (10 %% 3 == 0) {
  print("10 is divisible by 3")
  } else if (10 %% 2 == 0) {
    print("10 is divisible by 2")
}
```

10 being obviously divisible by 2 and not 3, it is the second phrase that will be printed. The
`%%` operator is the modulus operator, which gives the rest of the division of 10 by 2.
It is easier to use `dplyr::case_when()`:

```{r}
case_when(10 %% 3 == 0 ~ "10 is divisible by 3",
          10 %% 2 == 0 ~ "10 is divisible by 2")
```

We have already encountered this function in Chapter 4, inside a `dplyr::mutate()` call to create a new column.

Let's now discuss loops.

### For loops

For loops make it possible to repeat a set of instructions `i` times. For example, try the following:

```{r, cache=TRUE}
for (i in 1:10){
  print("hello")
}
```

It is also possible to do computations using for loops. Let's compute the sum of the first
100 integers:

```{r, cache=TRUE}
result = 0
for (i in 1:100){
  result = result + i
}

print(result)
```

`result` is equal to 5050, the expected result. What happened in that loop? First, we defined a
variable called `result` and set it to 0. Then, when the loops starts, `i` equals 1, so we add
`result` to `1`, which is 1. Then, `i` equals 2, and again, we add `result` to `i`. But this time,
`result` equals 1 and `i` equals 2, so now `result` equals 3, and we repeat this until `i`
equals 100. As explained in the introduction, this is not my preferred way of doing something like
that. I actually use loops very rarely. I would solve the above problem using `purrr::reduce()`:

```{r}
reduce(seq(1,100), `+`)
```

This might look more complex, but once you're used to it, you will not use many loops again.
You might wonder then, why use loops at all in this case? I refer you to the following
[section](http://adv-r.had.co.nz/Functionals.html#functionals-not) of Hadley Wickham's *Advanced R*
for an in-depth discussion on situations where loops make more sense than higher-order functions
such as `purrr::map()` or `purrr::reduce()`.

### While loops

While loops are very similar to for loops. The instructions inside a while loop are repeated while a
certain condition holds true. Let's consider the sum of the first 100 integers again:

```{r, cache=TRUE}
result = 0
i = 1
while (i<=100){
  result = result + i
  i = i + 1
}

print(result)
```

Here, we first set `result` and `i` to 0. Then, while `i` is less than, or equal to 100, we add `i`
to `result`. Notice that there is one more line than in the for loop: we need to increment the value
of `i`, if not, `i` would stay equal to 1, and the condition would always be fulfilled, and
the loop would run forever (not really, only until your computer runs out of memory, or until the
heat death of the universe, whichever comes first).

Now that we know how to write loops, and know about `if...else...` constructs, we have (almost) all
the ingredients to write our own functions.

## Writing your own functions

As you have seen by now, R includes a very large amount of preprogrammed functions, but also many
more functions are available in packages. However, there will be a lot of situations where you will
need to write your own. In this section we are going to learn how to write our own functions.

### Declaring functions in R

Suppose you want to create the following function: \(f(x) = \dfrac{1}{\sqrt{x}}\).
Writing this in R is quite simple:

```{r, cache=TRUE}
my_function <- function(x){
1/sqrt(x)
}
```

The argument of the function, `x`, gets passed to the... `function()` function and the *body* of 
the function (more on that in the next Chapter) contains the function definition. Of course, 
you could define functions that use more than one input:

```{r, cache=TRUE}
my_function <- function(x, y){
1/sqrt(x + y)
}
```

or inputs with names longer than one character:

```{r, cache=TRUE}
my_function <- function(argument1, argument2){
1/sqrt(argument1 + argument2)
}
```

Functions written by the user get called just the same way as functions included in R:

```{r, cache=TRUE}
my_function(1, 10)
```

It is also possible to provide default values to the function's arguments. Values that are used
if the user omits them:

```{r, cache=TRUE}
my_function <- function(argument1, argument2 = 10){
1/sqrt(argument1 + argument2)
}
```

```{r, cache=TRUE}
my_function(1)
```

This is especially useful for functions with many arguments, or to allow users to choose between
different methods:

```{r, cache=TRUE}
my_function <- function(argument1, argument2, method = "foo"){
  
  x <- argument1 + argument2
  
  if(method == "foo"){
    1/sqrt(x)
  } else if (method == "bar"){
    "this is a string"
    }
}

my_function(10, 11)

my_function(10, 11, "bar")
```

As you see, depending on the "method" chosen, the returned result is either a numeric, or a string.
What happens if the user provides a "method" that is neither "foo" nor "bar"?

```{r, cache=TRUE}
my_function(10, 11, "spam")
```

As you can see nothing happens. It is possible to add safeguards to your function to avoid such
situations:

```{r, cache=TRUE}
my_function <- function(argument1, argument2, method = "foo"){
  
  if(!(method %in% c("foo", "bar"))){
    return("Method must be either 'foo' or 'bar'")
  }
  
  x <- argument1 + argument2
  
  if(method == "foo"){
    1/sqrt(x)
  } else if (method == "bar"){
    "this is a string"
    }
}

my_function(10, 11)

my_function(10, 11, "bar")

my_function(10, 11, "foobar")
```

Notice that I have used `return()` inside my first `if` statement. This is to immediately stop
evaluation of the function and return a value. If I had omitted it, evaluation would have 
continued, as it is always the last expression that gets evaluated. Remove `return()` and run the
function again, and see what happens. Later, we are going to learn how to add better safeguards to 
your functions, to avoid runtime errors.

While in general, it is a good idea to add comments to your functions to explain what they do, I
would avoid adding comments to functions that do things that are very obvious, such as with this
one. Function names should be of the form: `function_name()`. Always give your function very
explicit names! In mathematics it is standard to give functions just one letter as a name, but I
would advise against doing that in your code. Functions that you write are not special in any way;
this means that R will treat them the same way, and they will work in conjunction with any other
function just as if it was built-in into R. 

They have one limitation though (which is shared with R's native function): just like in math, 
they can only return one value. However, sometimes, you may need to return more than one value. 
To be able to do this, you must put your values in a list, and return the list of values. For example:


```{r, cache=TRUE}
average_and_sd <- function(x){
c(mean(x), sd(x))
}

average_and_sd(c(1, 3, 8, 9, 10, 12))
```

You're still returning a single object, but it's a vector. You can also return a named list:

```{r, cache=TRUE}
average_and_sd <- function(x){
list("mean_x" =  mean(x), "sd_x" = sd(x))
}

average_and_sd(c(1, 3, 8, 9, 10, 12))
```

As described before, you can use `return()` at the end of your functions:

```{r, cache=TRUE}
average_and_sd <- function(x){
  result <- c(mean(x), sd(x))
return(result)
}

average_and_sd(c(1, 3, 8, 9, 10, 12))
```

But this is only needed if you need to return a value early:

```{r, cache=TRUE}
average_and_sd <- function(x){
if(any(is.na(x))){
    return(NA)
  } else {
    c(mean(x), sd(x))
    }
}

average_and_sd(c(1, 3, 8, 9, 10, 12))
average_and_sd(c(1, 3, NA, 9, 10, 12))
```

If you need to use a function from a package inside your function use `::`:

```{r}
my_sum <- function(a_vector){
  purrr::reduce(a_vector, `+`)
}
```

However, if you need to use more than one function, this can become tedious. A quick and dirty
(more dirty than quick, really) way of doing that, is to use `library(package_name)`, inside the function:

```{r}
my_sum <- function(a_vector){
  library(purrr)
  reduce(a_vector, `+`)
}
```

Loading the library inside the function has the advantage that you will be sure that the package
upon which your function depends will be loaded. If the package is already loaded, it will not be
loaded again, thus not impact performance, but if you forgot to load it at the beginning of your
script, then, no worries, your function will load it the first time you use it! However, the very
best way would be to write your own package and declare the packages upon which your functions
depend as dependencies. This is something we are going to explore in Chapter 11.

You can put a lot of instructions inside a function, such as loops. Let's create the function that
returns Fionacci numbers.

### Fibonacci numbers

The Fibonacci sequence is the following:

$$1, 1, 2, 3, 5, 8, 13, 21, 34, 55, ...$$

Each subsequent number is composed of the sum of the two preceding ones. In R, it is possible to define a function that returns the $n^{th}$ my_fibonacci number:

```{r}
my_fibo <- function(n){
 a <- 0
 b <- 1
 for (i in 1:n){
  temp <- b
  b <- a
  a <- a + temp
 }
 a
}
```

Inside the loop, we defined a variable called `temp`. Defining temporary variables is usually very
useful. Let's try to understand what happens inside this loop:

* First, we assign the value 0 to variable `a` and value 1 to variable `b`.
* We start a loop, that goes from 1 to `n`.
* We assign the value inside of `b` to a temporary variable, called `temp`.
* `b` becomes `a`.
* We assign the sum of `a` and `temp` to `a`.
* When the loop is finished, we return `a`.

What happens if we want the 3rd my_fibonacci number? At `n = 1` we have first `a = 0` and `b = 1`,
then `temp = 1`, `b = 0` and `a = 0 + 1`. Then `n = 2`. Now `b = 0` and `temp = 0`. The previous
result, `a = 0 + 1` is now assigned to `b`, so `b = 1`. Then, `a = 1 + 0`. Finally, `n = 3`. `temp
= 1` (because `b = 1`), the previous result `a = 1` is assigned to `b` and finally, `a = 1 + 1`. So
the third my_fibonacci number equals 2. Reading this might be a bit confusing; I strongly advise you
to run the algorithm on a sheet of paper, step by step.

The above algorithm is called an iterative algorithm, because it uses a loop to compute the result.
Let's look at another way to think about the problem, with a so-called recursive function:

```{r, cache=TRUE}
fibo_recur <- function(n){
 if (n == 0 || n == 1){
   return(n)
   } else {
   fibo_recur(n-1) + fibo_recur(n-2)
   }
}
```

This algorithm should be easier to understand: if `n = 0` or `n = 1` the function should return `n`
(0 or 1). If `n` is strictly bigger than `1`, `fibo_recur()` should return the sum of
`fibo_recur(n-1)` and `fibo_recur(n-2)`. This version of the function is very much the same as the
mathematical definition of the fibonacci sequence. So why not use only recursive algorithms
then? Try to run the following:

```{r, cache=TRUE}
system.time(my_fibo(30))
```

The result should be printed very fast (the `system.time()` function returns the time that it took
to execute `my_fibo(30)`). Let's try with the recursive version:

```{r, cache=TRUE}
system.time(fibo_recur(30))
```

It takes much longer to execute! Recursive algorithms are very CPU demanding, so if speed is
critical, it's best to avoid recursive algorithms. Also, in `fibo_recur()` try to remove this line:
`if (n == 0 || n == 1)` and try to run `fibo_recur(5)` for example and see what happens. You should
get an error: this is because for recursive algorithms you need a stopping condition, or else,
it would run forever. This is not the case for iterative algorithms, because the stopping
condition is the last step of the loop.

So as you can see, for recursive relationships, for or while loops are the way to go in R, whether
you're writing these loops inside functions or not.

## Exercises

### Exercise 1 {-}

In this exercise, you will write a function to compute the sum of the n first integers. Combine the
algorithm we saw in section about while loops and what you learned about functions
in this section.

```{r, include=FALSE}
MySum <- function(n){
  result = 0
  i = 1
  while (i<=n){
    result = result + i
    i = i + 1
  }
  result
}
```

### Exercise 2 {-}

Write a function called `my_fact()` that computes the factorial of a number `n`. Do it using a
loop, using a recursive function, and using a functional:

```{r, include=FALSE}
my_fact_iter <- function(n){
  result = 1
  for(i in 1:n){
    result = result * i
    i = i + 1
  }
  result
}

my_fact_recur <- function(n){
  if(n == 0 || n == 1){
  result = 1
  } else {
    n * MyFactorialRecur(n-1)
  }
}

my_fact_reduce <- function(n){
  reduce(seq(1, n), `*`)
}
```

### Exercise 3 {-}

Write a function to find the roots of quadratic functions. Your function should take 3 arguments,
`a`, `b` and `c` and return the two roots. Only consider the case where there are two real roots
(delta > 0).

```{r, include=FALSE}
quad_root <- function(a, b, c){
# function that returns the root of a quadratic function
# very basic, doesn't cover the case where delta < 0
  delta = b**2 - 4 * a * c
  x1 = (-b + sqrt(delta)) / (2 * a)
  x2 = (-b - sqrt(delta)) / (2 * a)
  c(x1, x2)
}

quad_root(1, -4, 3) # should return 3 and 1
```

## Functions that take functions as arguments: writing your own higher-order functions

Functions that take functions as arguments are very powerful and useful tools. You already know a
couple, `purrr::map()` and `purrr::reduce()`, discussed briefly in Chapter 4.
But you can also write your own! A very simple example would be the following:

```{r}
my_func <- function(x, func){
  func(x)
}
```

`my_func()` is a very simple function that takes `x` and `func()` as arguments and that simply
executes `func(x)`. This might not seem very useful (after all, you could simply use `func(x)!`) but
this is just for illustration purposes, in practice, your functions would be more useful than that!
Let's try to use `my_func()`:

```{r}
my_func(c(1, 8, 1, 0, 8), mean)
```

As expected, this returns the mean of the given vector. But now suppose the following:

```{r}
my_func(c(1, 8, 1, NA, 8), mean)
```

Because one element of the list is `NA`, the whole mean is `NA`. `mean()` has a `na.rm` argument
that you can set to `TRUE` to ignore the `NA`s in the vector. However, here, there is no way to
provide this argument to the function `mean()`! Let's see what happens when we try to:

```{r, eval=FALSE}
my_func(c(1, 8, 1, NA, 8), mean, na.rm = TRUE)
```

```
Error in my_func(c(1, 8, 1, NA, 8), mean, na.rm = TRUE) :
  unused argument (na.rm = TRUE)
```

So what you could do is pass the value `TRUE` to the `na.rm` argument of `mean()` from your own
function:

```{r}
my_func <- function(x, func, remove_na){
  func(x, na.rm = remove_na)
}

my_func(c(1, 8, 1, NA, 8), mean, remove_na = TRUE)
```

This is one solution, but `mean()` also has another argument called `trim`. What if some other
user needs this argument? Should you also add it to your function? Surely there's a way to avoid
this problem? Yes, there is, and it's the *dots*. The `...` simply mean "any other
argument as needed", and it's very easy to use:

```{r}
my_func <- function(x, func, ...){
  func(x, ...)
}

my_func(c(1, 8, 1, NA, 8), mean, na.rm = TRUE)
```

or, now, if you need the `trim` argument:

```{r}
my_func(c(1, 8, 1, NA, 8), mean, na.rm = TRUE, trim = 0.1)
```

The `...` are very useful when writing wrappers such as `my_func()`.

## Functions that take columns of data as arguments

In many situations, you will want to write functions that look similar to this:

```{r, eval=FALSE}
my_function(my_data, one_column_inside_data)
```

Such a function would be useful in situation where you have to apply a certain number of operations
to columns for different data frames. For example if you need to create tables of descriptive
statistics or graphs periodically, it might be very interesting to put these operations inside a
function and then call the function whenever you need it, on the fresh batch of data.

However, if you try to write something like that, something that might seem unexpected, at first,
will happen:

```{r, eval=FALSE}
data(mtcars)

simple_function <- function(dataset, col_name){
  dataset %>%
    group_by(col_name) %>%
    summarise(mean_speed = mean(speed))
}


simple_function(cars, "dist")
```

```
Error: unknown variable to group by : col_name
```

The variable `col_name` is passed to `simple_function()` as a string, but `group_by()` requires a
variable name. So why not try to convert `col_name` to a name?

```{r, eval=FALSE}
simple_function <- function(dataset, col_name){
  col_name <- as.name(col_name)
  dataset %>%
    group_by(col_name) %>%
    summarise(mean_speed = mean(speed))
}


simple_function(cars, "dist")
```

```
Error: unknown variable to group by : col_name
```

This is because R is literally looking for the variable `"dist"` somewhere in the global
environment, and not as a column of the data. R does not understand that you are refering to the
column `"dist"` that is inside the dataset. So how can we make R understands what you mean?

To be able to do that, we need to use a framework that was introduced recently in the `{tidyverse}`,
called `{tidyeval}`. This discussion can get very technical, so I will spare you the details.
However, you can read about it [here](http://dplyr.tidyverse.org/articles/programming.html) and
[here](https://cran.r-project.org/web/packages/rlang/vignettes/tidy-evaluation.html). The
discussion can get complicated, but using `{tidyeval}` is actually quite easy, and you can get a
cookbook approach to it. Take a look at the code below:


```{r}
simple_function <- function(dataset, col_name){
  col_name <- enquo(col_name)
  dataset %>%
    group_by(!!col_name) %>%
    summarise(mean_mpg = mean(mpg))
}


simple_function(mtcars, cyl)
```

As you can see, the previous idea we had, which was using `as.name()` was not very far away from
the solution. The solution, with `tidyeval`, consists in using `enquo()`, which for our purposes,
let's say that this function does something similar to `as.name()` (in truth, it doesn't). Now that
`col_name` is (R programmers call it) quoted, we need to tell `group_by()` to evaluate the input as
is. This is done with `!!()`, which is another `{tidyeval}` function. I say it again;
don't worry if you don't understand everything. Just remember to use `enquo()` on your column names
and then `!!()` inside the `{dplyr}` function you want to use.

Let's see some other examples:

```{r}
simple_function <- function(dataset, col_name, value){
  col_name <- enquo(col_name)
  dataset %>%
    filter((!!col_name) == value) %>%
    summarise(mean_cyl = mean(cyl))
}


simple_function(mtcars, am, 1)
```

Notice that I’ve written:

```{r, eval=FALSE}
filter((!!col_name) == value)
```
and not:

```{r, eval=FALSE}
filter(!!col_name == value)
```
I have enclosed `!!col_name` inside parentheses. This is because operators such as `==` have
precedence over `!!`, so you have to be explicit. Also, notice that I didn't have to quote `1`.
This is because it's *standard* variable, not a column inside the dataset. Let’s make this function
a bit more general. I hard-coded the variable cyl inside the body of the function, but maybe you’d
like the mean of another variable?

```{r}
simple_function <- function(dataset, filter_col, mean_col, value){
  filter_col <- enquo(filter_col)
  mean_col <- enquo(mean_col)
  dataset %>%
    filter((!!filter_col) == value) %>%
    summarise(mean((!!mean_col)))
}


simple_function(mtcars, am, cyl, 1)
```

Notice that I had to quote `mean_col` too.

Using the `...` that we discovered in the previous section, we can pass more than one column:

```{r}
simple_function <- function(dataset, ...){
  col_vars <- quos(...)
  dataset %>%
    summarise_at(vars(!!!col_vars), funs(mean, sd))
}
```

Because these *dots* contain more than one variable, you have to use `quos()` instead of `enquo()`.
This will put the arguments provided via the dots in a list. Then, because we have a list of
columns, we have to use `summarise_at()`, which you should know if you did the exercices of
Chapter 4. So if you didn't do them, go back to them and finish them first. Doing the exercise will
also teach you what `vars()` and `funs()` are. The last thing you have to pay attention to is to
use `!!!()` if you used `quos()`. So 3 `!` instead of only 2. This allows you to then do things
like this:

```{r}
simple_function(mtcars, am, cyl, mpg)
```

Using `...` with `!!!()` allows you to write very flexible functions.

If you need to be even more general, you can also provide the summary functions as arguments of
your function, but you have to rewrite your function a little bit:

```{r}
simple_function <- function(dataset, cols, funcs){
  dataset %>%
    summarise_at(vars(!!!cols), funs(!!!funcs))
}
```

You might be wondering where the `quos()` went? Well because now we are passing two lists of, a list
columns that we have to quote, and a list of functions, that we have to quote, we need to use `quos()`
when calling the function:

```{r}
simple_function(mtcars, quos(am, cyl, mpg), quos(mean, sd, sum))
```

This works, but I don't think you'll need to have that much flexibility; either the columns
are variable, or the functions, but rarely both at the same time.

## Functions that use loops

It is entirely possible to put a loop inside a function. For example, consider the following
function that return the square root of a number using Newton's algorithm:

```{r}
sqrt_newton <- function(a, init = 1, eps = 0.01){
    stopifnot(a >= 0)
    while(abs(init**2 - a) > eps){
        init <- 1/2 *(init + a/init)
    }
    init
}
```

This functions contains a while loop inside its body. Let's see if it works:

```{r}
sqrt_newton(16)
```

In the definition of the function, I wrote `init = 1` and `eps = 0.01` which means that this
argument can be omitted and will have the provided value (0.01)  as the default. You can then use
this function as any other, for example with `map()`:

```{r}
map(c(16, 7, 8, 9, 12), sqrt_newton)
```

This is what I meant before with "your functions are nothing special". Once the function is
defined, you can use it like any other base R function.

Notice the use of `stopifnot()` inside the body of the function. This is a way to return an error
in case a condition is not fulfilled. We are going to learn more about this type of functions
in the next chapter.

## Anonymous functions

As the name implies, anonymous functions are functions that do not have a name. These are useful inside
functions that have functions as arguments, such as `purrr::map()` or `purrr::reduce()`:

```{r}
map(c(1,2,3,4), function(x){1/sqrt(x)})
```

These anonymous functions get defined in a very similar way to regular functions, you just skip the
name and that's it. `tidyverse` functions also support formulas; these get converted to anonymous functions:

```{r}
map(c(1,2,3,4), ~{1/sqrt(.)})
```

Using a formula instead of an anonymous function is less verbose; you use `~` instead of `function(x)`
and a single dot `.` instead of `x`. What if you need an anonymous function that requires more than
one argument? This is not a problem:

```{r}
map2(c(1, 2, 3, 4, 5), c(9, 8, 7, 6, 5), function(x, y){(x**2)/y})
```

or, using a formula:

```{r}
map2(c(1, 2, 3, 4, 5), c(9, 8, 7, 6, 5), ~{(.x**2)/.y})
```

Because you have now two arguments, a single dot could not work, so instead you use `.x` and `.y` to
avoid confusion.

You now know a lot about writing your own functions. In the next chapter, we are going to learn about
functional programming, the programming paradigm I described in the introduction of this book.

## Exercises

### Exercise 1 {-}

* Create the following vector:

\[a = (1,6,7,8,8,9,2)\]

Using a for loop and a while loop, compute the sum of its elements. To avoid issues, use `i`
as the counter inside the for loop, and `j` as the counter for the while loop.

```{r, include=FALSE}
result = 0
  for(i in 1:length(a)){
    result = result + a[i]
}

result = 0
  while (i <= length(a)){
    result = result + a[i]
    i = i + 1
}
```

* How would you achieve that with a functional (a function that takes a function as an argument)?

### Exercise 2 {-}

* Let's use a loop to get the matrix product of a matrix A and B. Follow these steps to create the loop:

1) Create matrix A:

\[A = \left(
  \begin{array}{ccc}
   9 & 4 & 12 \\
   5 & 0 & 7 \\
   2 & 6 & 8 \\
   9 & 2 & 9
  \end{array} \right)
\]

2) Create matrix B:

\[B = \left(
\begin{array}{cccc}
 5 & 4 & 2 & 5 \\
 2 & 7 & 2 & 1 \\
 8 & 3 & 2 & 6 \\
\end{array} \right)
\]

3) Create a matrix C, with dimension 4x4 that will hold the result. Use this command: `C = matrix(rep(0,16), nrow = 4)}

4) Using a for loop, loop over the rows of A first: `for(i in 1:nrow(A))}

5) Inside this loop, loop over the columns of B: `for(j in 1:ncol(B))}

6) Again, inside this loop, loop over the rows of B: `for(k in 1:nrow(B))}

7) Inside this last loop, compute the result and save it inside C: `C[i,j] = C[i,j] + A[i,k] * B[k,j]}

8) Now write a function that takes two matrices as arguments, and returns their product.

```{r, include=FALSE}
A <- matrix( c(9, 4, 12, 5, 0, 7, 2, 6, 8, 9, 2, 9), nrow = 4, byrow = TRUE)

B <- matrix( c(5, 4, 2, 5, 2, 7, 2, 1, 8, 3, 2, 6), nrow = 3, byrow = TRUE)

C <- matrix(rep(0,16), nrow = 4)

for(i in 1:nrow(A)){
  for(j in 1:ncol(B)){
    for(k in 1:nrow(B)){
      C[i,j] = C[i,j] + A[i,k] * B[k,j]
    }
  }
}
```

* R has a built-in function to compute the dot product of 2 matrices. Which is it?

### Exercise 3 {-}

* Fizz Buzz: Print integers from 1 to 100. If a number is divisible by 3, print the word `"Fizz"` if
it's divisible by 5, print `"Buzz"`. Use a for loop and if statements.

* Write a function that takes an integer as arguments, and prints `"Fizz"` or `"Buzz"` up to that integer.

### Exercise 4 {-}

* Fizz Buzz 2: Same as above, but now add this third condition: if a number is both divisible by 3 and 5, print `"FizzBuzz"`.

```{r, include=FALSE}
for (i in 1:100){
if (i %% 15 == 0) {
  print("FizzBuzz")
} else if (i %% 3 == 0) {
  print("Fizz")
} else if (i %% 5 == 0) {
  print("Buzz")
} else {
  print(i)
}
}
```

* Write a function that takes an integer as argument, and prints `Fizz`, `Buzz` or `FizzBuzz` up to that integer.
